%==============================================================
%  Acknowledgments Page
%==============================================================

\cleardoublepage
\thispagestyle{empty}

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

\setstretch{1.4}

I would like to begin by expressing my sincere gratitude to everyone who contributed, directly or indirectly, to the completion of this final year project.

\vspace{5pt}

First and foremost, I would like to thank my supervisor, \textbf{Mr.~Othmani Mohamed}, for his valuable guidance, constructive feedback, and continuous availability throughout the development of this work. His advice and encouragement played a key role in shaping the direction of this project.

\vspace{5pt}

I also wish to extend my appreciation to all the teachers of the \textbf{Faculty of Sciences of Gafsa}, who shared their knowledge and experience with us during these years of study. Their dedication helped build the technical foundation that made a project of this scale possible.

\vspace{5pt}

A special thanks goes to the members of the jury, who kindly accepted to evaluate this work. I am honored by their time and attention.

\vspace{5pt}

I am deeply grateful to my parents and family for their unconditional support, patience, and constant encouragement throughout my entire academic journey. Their belief in me has been the strongest motivation.

\vspace{5pt}

Finally, I would like to thank my friends and classmates for the moments of collaboration, discussion, and mutual support that made this experience much more meaningful.

\vspace{10pt}

\begin{flushright}
\textit{Mohamed Mbarek}
\end{flushright}
